---
title: "Distance Analysis"
author: "Owen Travis"
date: "Princeton University Senior Thesis, Class of 2024"
output:
  html_document:
    df_print: paged
    toc: no
    toc_depth: 3
  pdf_document:
    fig_caption: yes
    number_sections: no
    toc: no
    toc_depth: 3
geometry: margin=1.5in
---

This file contains code used for processing the output of `distance.py`.
This provides an initial understanding of the distance between each move and
the corresponding previous move, without any AI analysis of positions.

### Setup
```{r, message=FALSE}
# ------------------------------------------------------------------------------
# Import necessary libraries
# ------------------------------------------------------------------------------
library(readr)
library(ggplot2)
library(magrittr)
library(dplyr)
library(forcats)
library(fBasics)
library(scales)
library(stringr)
library(gridExtra)
```

```{r, message=FALSE}
# ------------------------------------------------------------------------------
# Load the data set
# ------------------------------------------------------------------------------
df = read_csv("/path/to/distance.py/output.csv")
```

```{r}
# ------------------------------------------------------------------------------
# Update column types and add distance columns
# ------------------------------------------------------------------------------
# Make isAlphaGo a factor
df$isAlphaGo <- as.factor(df$isAlphaGo)
# Set the AlphaGo player rank to 'AG'
df[df$isAlphaGo==1, "playerRank"] <- 'AG'
# Make player rank a factor
df$Rank <- as.factor(df$playerRank)
# Create a custom ordering of player rank
df <- df %>% mutate(orderedRank = fct_relevel(playerRank, "30k", "29k", "28k",
                "27k", "26k", "25k", "24k", "23k", "22k", "21k", "20k", "19k", "18k",
                '17k', '16k', '15k', '14k', "13k", "12k", '11k', "10k", "9k", "8k", "7k",
                "6k",'5k', '4k', '3k', "2k", '1k', "1d", "2d", "3d", "4d", "5d", "6d", "7d",
                "8d", "9d", "AG"))
# Calculate city block (sum) distance and Euclidean distance from the previous move
df <- df %>% mutate(sum_played_dx_dy = played_dx + played_dy)
df <- df %>% mutate(euclidean_played_dx_dy = sqrt(played_dx^2 + played_dy^2))
```

```{r}
# ------------------------------------------------------------------------------
# Flag additional robots
# ------------------------------------------------------------------------------
df[str_detect(tolower(df$playerName), 'alphago'), "isBot"] <- TRUE
df[str_detect(tolower(df$playerName), 'leelazero'), "isBot"] <- TRUE
```

### Game and move counts
```{r}
# ------------------------------------------------------------------------------
# Report counts
# ------------------------------------------------------------------------------
 #AlphaGo game count
df %>% filter(isAlphaGo == 1) %>% select(gameFile) %>% unique() %>% nrow()
#AlphaGo move count
df %>% filter(isAlphaGo == 1) %>% nrow()
#KGS game count
df %>% filter(isAlphaGo == 0) %>% select(gameFile) %>% unique() %>% nrow()
#KGS move count
df %>% filter(isAlphaGo == 0) %>% nrow()
#KGS human move count
df %>% filter(isAlphaGo == 0 & isBot == FALSE) %>% nrow()
#KGS robot move count
df %>% filter(isAlphaGo == 0 & isBot == TRUE) %>% nrow()
```

### Move distribution by player rank
```{r}
# ------------------------------------------------------------------------------
# Figure: Distribution of human moves by player rank
# ------------------------------------------------------------------------------
df %>% filter(isAlphaGo == 0 & isBot == FALSE) %>% 
  filter(orderedRank %in% c("6k",'5k', '4k', '3k', "2k", '1k', "1d", "2d", "3d", "4d", "5d",
                            "6d", "7d", "8d", "9d")) %>% 
  ggplot(aes(x=orderedRank)) +
  geom_bar() +
  labs(x=expression("Increasing player rank" %->% ""),
       y="Number of moves observed") +
  scale_y_continuous(labels = scales::label_number_si()) +
  theme(axis.title=element_text(size=20),
    axis.text.x=element_text(size=20),
    axis.text.y=element_text(size=20),
    legend.text=element_text(size=20))
```

### Density plots of move distances
```{r}
# ------------------------------------------------------------------------------
# Figure: Density plot of city block (sum) distance for AlphaGo and humans
# ------------------------------------------------------------------------------
df %>%
  filter(isAlphaGo == 1 | isBot == FALSE) %>% #select only AlphaGo and human moves
  ggplot() +
  geom_density(aes(x=sum_played_dx_dy, fill=isAlphaGo), alpha=0.5, na.rm=T, bw=1) +
  labs(x="City block distance from the previous move", y="Density", fill="") + 
  scale_fill_discrete(labels=c("Human", "AlphaGo")) +
  theme(axis.title=element_text(size=20),
      axis.text.x=element_text(size=20),
      axis.text.y=element_text(size=20),
      legend.text=element_text(size=20))
```

```{r}
# ------------------------------------------------------------------------------
# Figure: Density plot using euclidean distance
# ------------------------------------------------------------------------------
df %>%
  filter(isAlphaGo == 1 | isBot == FALSE) %>% 
  ggplot() +
  geom_density(aes(x=euclidean_played_dx_dy, fill=isAlphaGo), alpha=0.5, na.rm=T, bw=1) +
  labs(x="Euclidean distance to the previous move", y="Density", fill="") + 
  scale_fill_discrete(labels=c("Human", "AlphaGo")) +
  theme(axis.title=element_text(size=20),
      axis.text.x=element_text(size=20),
      axis.text.y=element_text(size=20),
      legend.text=element_text(size=20))

alphago_quantile <- df %>% filter(isAlphaGo == 1)
alphago_quantile$sum_played_dx_dy %>% quantile(.50, na.rm=TRUE)
alphago_quantile %>% filter(sum_played_dx_dy <= 3) %>% nrow()
alphago_quantile %>% nrow()
```

### Bar plots of average move distances
```{r}
# ------------------------------------------------------------------------------
# Figure: Bar plot of averages using city block distance
# ------------------------------------------------------------------------------
# Pull the list of human city block (sum) move distances
humanSumDistances <- df %>%
  filter(isAlphaGo == 0 & isBot == FALSE) %>%
  pull(sum_played_dx_dy) %>% 
  na.omit()
# Calculate the mean, SEM, and 95% confidence interval
humAvgSumDist <- mean(humanSumDistances)
humSumSEM <- sd(humanSumDistances) / sqrt(length(humanSumDistances))
humBoundSumDiff <- qnorm(0.975)*humSumSEM

# Pull the list of KGS robot city block (sum) move distances
otherBotSumDistances <- df %>% 
  filter(isAlphaGo == 0 & isBot == TRUE) %>% 
  pull(sum_played_dx_dy) %>% 
  na.omit()
# Calculate the mean, SEM, and 95% confidence interval
otherBotAvgSumDist <- mean(otherBotSumDistances)
otherBotSumSEM <- sd(otherBotSumDistances) / sqrt(length(otherBotSumDistances))
otherBotBoundSumDiff <- qnorm(0.975)*otherBotSumSEM

# Pull the list of AlphaGo city block (sum) move distances
alphaGoSumDistances <- df %>%
  filter(isAlphaGo == 1) %>% 
  pull(sum_played_dx_dy) %>% 
  na.omit()
# Calculate the mean, SEM, and 95% confidence interval
alphaGoAvgSumDist <- mean(alphaGoSumDistances)
alphaGoSumSEM <- sd(alphaGoSumDistances) / sqrt(length(alphaGoSumDistances))
alphaGoBoundSumDiff <- qnorm(0.975)*alphaGoSumSEM

# Create a data frame with means and confidence intervals
data_fig_mean_sum_distance <- data.frame(name=fct_relevel(factor(c("Human", "Robot on KGS", "AlphaGo")),
                                                          "Human", "Robot on KGS", "AlphaGo"),
                        value=c(humAvgSumDist, otherBotAvgSumDist, alphaGoAvgSumDist),
                        ci=c(humBoundSumDiff, otherBotBoundSumDiff, alphaGoBoundSumDiff))

# Make the plot
fig_mean_sum_distance <- ggplot(data_fig_mean_sum_distance, aes(x=name, y=value, fill=name)) +
  geom_bar(stat="identity", show.legend=F) + 
  geom_errorbar(aes(x=name, ymin=value-ci, ymax=value+ci, colour=name), size=1, width=0.4) +
  labs(fill="", x="", y="Mean Manhattan distance to the previous move") +
  theme(axis.title=element_text(size=30),
        axis.text.x=element_text(size=25),
        axis.text.y=element_text(size=30),
        legend.position="none") + 
  scale_fill_manual(values=c("#F8766D", "#00BA38", "#00BFC4")) +
  scale_color_manual(values=c("#F8766D", "#00BA38", "#00BFC4"))
```

```{r}
# ------------------------------------------------------------------------------
# Figure: Bar plot of averages using euclidean distance
# ------------------------------------------------------------------------------
# Pull the list of human Euclidean move distances
humanEucDistances <- df %>%
  filter(isAlphaGo == 0 & isBot == FALSE) %>%
  pull(euclidean_played_dx_dy) %>% 
  na.omit()
# Calculate the mean, SEM, and 95% confidence interval
humAvgEucDist <- mean(humanEucDistances)
humEucSEM <- sd(humanEucDistances) / sqrt(length(humanEucDistances))
humBoundEucDiff <- qnorm(0.975)*humEucSEM

# Pull the list of KGS robot Euclidean move distances
otherBotEucDistances <- df %>% 
  filter(isAlphaGo == 0 & isBot == TRUE) %>% 
  pull(euclidean_played_dx_dy) %>% 
  na.omit()
# Calculate the mean, SEM, and 95% confidence interval
otherBotAvgEucDist <- mean(otherBotEucDistances)
otherBotEucSEM <- sd(otherBotEucDistances) / sqrt(length(otherBotEucDistances))
otherBotBoundEucDiff <- qnorm(0.975)*otherBotEucSEM

# Pull the list of AlphaGo Euclidean move distances
alphaGoEucDistances <- df %>%
  filter(isAlphaGo == 1) %>% 
  pull(euclidean_played_dx_dy) %>% 
  na.omit()
# Calculate the mean, SEM, and 95% confidence interval
alphaGoAvgEucDist <- mean(alphaGoEucDistances)
alphaGoEucSEM <- sd(alphaGoEucDistances) / sqrt(length(alphaGoEucDistances))
alphaGoBoundEucDiff <- qnorm(0.975)*alphaGoEucSEM

# Create a data frame with means and confidence intervals
data_fig_mean_euc_distance <- data.frame(name=fct_relevel(factor(c("Human", "Robot on KGS", "AlphaGo")),
                                                     "Human", "Robot on KGS", "AlphaGo"),
                        value=c(humAvgEucDist, otherBotAvgEucDist, alphaGoAvgEucDist),
                        ci=c(humBoundEucDiff, otherBotBoundEucDiff, alphaGoBoundEucDiff))

fig_mean_euc_distance <- ggplot(data_fig_mean_euc_distance, aes(x=name, y=value, fill=name)) +
  geom_bar(stat="identity", show.legend=F) + 
  geom_errorbar(aes(x=name, ymin=value-ci, ymax=value+ci, colour=name), size=1, width=0.4) +
  labs(fill="", x="", y="Mean Euclidean distance to the previous move") +
  theme(axis.title=element_text(size=30),
        axis.text.x=element_text(size=25),
        axis.text.y=element_text(size=30),
        legend.position="none") + 
  scale_fill_manual(values=c("#F8766D", "#00BA38", "#00BFC4")) +
  scale_color_manual(values=c("#F8766D", "#00BA38", "#00BFC4"))
```

```{r}
# ------------------------------------------------------------------------------
# Figure: Joint figure showing means in city block and Euclidean distance
# ------------------------------------------------------------------------------
grid.arrange(fig_mean_euc_distance, fig_mean_sum_distance, ncol=2)
```

### Statistics
```{r}
# ------------------------------------------------------------------------------
# Statistics for Euclidean distance
# ------------------------------------------------------------------------------
# Counts
length(humanEucDistances)
length(otherBotEucDistances)
length(alphaGoEucDistances)

# Averages
humAvgEucDist
otherBotAvgEucDist
alphaGoAvgEucDist

# Standard deviations
sd(humanEucDistances)
sd(otherBotEucDistances)
sd(alphaGoEucDistances)

# Two sample t-tests
t.test(x=humanEucDistances, y=alphaGoEucDistances, alternative="two.sided", mu=0)
t.test(x=humanEucDistances, y=otherBotEucDistances, alternative="two.sided", mu=0)
t.test(x=otherBotEucDistances, y=alphaGoEucDistances, alternative="two.sided", mu=0)

# Kolmogorov-Smirnov tests
ks.test(humanEucDistances, alphaGoEucDistances, alternative="two.sided")
ks.test(humanEucDistances, otherBotEucDistances, alternative="two.sided")
ks.test(otherBotEucDistances, alphaGoEucDistances, alternative="two.sided")
```

```{r}
# ------------------------------------------------------------------------------
# Statistics for city block (sum) distance
# ------------------------------------------------------------------------------
# Counts (evidently the same for Euclidean distance)
length(humanSumDistances)
length(otherBotSumDistances)
length(alphaGoSumDistances)

# Averages
humAvgSumDist
otherBotAvgSumDist
alphaGoAvgSumDist

# Standard deviations
sd(humanSumDistances)
sd(otherBotSumDistances)
sd(alphaGoSumDistances)

# Two sample t-tests
t.test(x=humanSumDistances, y=alphaGoSumDistances, alternative="two.sided", mu=0)
t.test(x=humanSumDistances, y=otherBotSumDistances, alternative="two.sided", mu=0)
t.test(x=otherBotSumDistances, y=alphaGoSumDistances, alternative="two.sided", mu=0)

# Kolmogorov-Smirnov tests
ks.test(humanSumDistances, alphaGoSumDistances, alternative="two.sided")
ks.test(humanSumDistances, otherBotSumDistances, alternative="two.sided")
ks.test(otherBotSumDistances, alphaGoSumDistances, alternative="two.sided")
```

### Quick visualizations of cumulative distribution functions
```{r}
# ------------------------------------------------------------------------------
# ECDF of human move distances vs. ECDF of AlphaGo move distances.
# ------------------------------------------------------------------------------
plot(ecdf(humanSumDistances), 
     xlim = range(c(humanSumDistances, alphaGoSumDistances)), 
     col = "blue",
     main="Human vs. AlphaGo")
plot(ecdf(alphaGoSumDistances), 
     add = TRUE, 
     lty = "dashed",
     col = "red")
```

```{r}
# ------------------------------------------------------------------------------
# ECDF of human move distances vs. ECDF of KGS robot move distances.
# ------------------------------------------------------------------------------
plot(ecdf(humanSumDistances), 
     xlim = range(c(humanSumDistances, otherBotSumDistances)), 
     col = "blue",
     main="Human vs. Other Robots on KGS")
plot(ecdf(otherBotSumDistances), 
     add = TRUE, 
     lty = "dashed",
     col = "red")
```

```{r}
# ------------------------------------------------------------------------------
# ECDF of KGS robot move distances vs. ECDF of AlphaGo move distances.
# ------------------------------------------------------------------------------
plot(ecdf(otherBotSumDistances), 
     xlim = range(c(otherBotSumDistances, alphaGoSumDistances)), 
     col = "blue",
     main="Human vs. Other Robots on KGS")
plot(ecdf(alphaGoSumDistances), 
     add = TRUE, 
     lty = "dashed",
     col = "red")
```

### Mean Euclidean move distance by player rank
```{r}
# ------------------------------------------------------------------------------
# Figure: Bar plot of averages by player rank using Euclidean distance
# ------------------------------------------------------------------------------
data_fig_mean_euc_distance_by_player_rank <- df %>%
  # keep only moves by AlphaGo or humans
  filter(isAlphaGo == 1 | isBot == FALSE) %>% 
  # keep only the higher ranks
  filter(orderedRank %in% c("6k", "5k", "4k", "3k", "2k", "1k", "1d", "2d", "3d",
                            "4d", "5d", "6d", "7d", "8d", "9d", "AG")) %>% 
  select(orderedRank, euclidean_played_dx_dy) %>% 
  na.omit() %>% 
  group_by(orderedRank) %>% 
  summarize(meanDist = mean(euclidean_played_dx_dy),
            ci = qnorm(0.975)*(sd(euclidean_played_dx_dy)/sqrt(length(euclidean_played_dx_dy))))
  
data_fig_mean_euc_distance_by_player_rank %>% 
  ggplot(aes(x=orderedRank, y=meanDist,
             # highlight the AlphaGo group
             fill=factor(ifelse(orderedRank=="AG","Normal","Highlighted")))) +
  geom_bar(stat="identity", show.legend=F) +
  geom_errorbar(aes(x=orderedRank, ymin=meanDist-ci,ymax=meanDist+ci,
                    # highlight the AlphaGo group
                    color=factor(ifelse(orderedRank=="AG","Normal","Highlighted"))),
                size=0.5, width=0.4) +
  labs(x=expression("Increasing player rank" %->% ""),
       y="Mean Euclidean distance to the previous move") +
  theme(axis.title=element_text(size=20),
        axis.text.x=element_text(size=20),
        axis.text.y=element_text(size=20),
        legend.position="none")
```

### Statistics
```{r}
# ------------------------------------------------------------------------------
# Correlation between average Euclidean distance to the previous move and
# player rank.
# ------------------------------------------------------------------------------
data_cor <- df %>%
  # keep only moves by AlphaGo or humans
  filter(isAlphaGo == 0 & isBot == FALSE) %>% 
  # keep only the higher ranks
  filter(orderedRank %in% c("6k", "5k", "4k", "3k", "2k", "1k", "1d", "2d", "3d",
                            "4d", "5d", "6d", "7d", "8d", "9d")) %>% 
  select(orderedRank, euclidean_played_dx_dy) %>% 
  na.omit() %>% 
  group_by(orderedRank) %>% 
  summarize(meanDist = mean(euclidean_played_dx_dy),
            ci = qnorm(0.975)*(sd(euclidean_played_dx_dy)/sqrt(length(euclidean_played_dx_dy))))

cor.test(c(-6, -5, -4, -3, -2, -1, 1, 2, 3, 4, 5, 6, 7, 8, 9),
         data_cor$meanDist, method=c("pearson"))

# ------------------------------------------------------------------------------
# Statistics for 9d players using Euclidean distance
# ------------------------------------------------------------------------------
# Pull the list of 9d human Euclidean move distances
human9dEucDistances <- df %>%
  filter(isAlphaGo == 0 & isBot == FALSE & playerRank == "9d") %>%
  pull(euclidean_played_dx_dy) %>% 
  na.omit()

# Count, mean, sd
length(human9dEucDistances)
mean(human9dEucDistances)
sd(human9dEucDistances)

# Two sample t-test on 9d distances and AlphaGo distances
t.test(x=human9dEucDistances, y=alphaGoEucDistances, alternative="two.sided", mu=0)
# Kolmogorov-Smirnov test on 9d distances and AlphaGo distances
ks.test(human9dEucDistances, alphaGoEucDistances, alternative="two.sided")
```

### Exploring the affect of move number on distance from the previous move
```{r}
# ------------------------------------------------------------------------------
# Figure: Mean Euclidean move distance by move number
#   (grouped by 10-percentiles for move numbers 1-250)
# ------------------------------------------------------------------------------
df %>%
  filter(num < 250 & num > 1) %>% 
  mutate(bin = ntile(num, n=10)) %>% 
  group_by(bin, isAlphaGo) %>% 
  summarize(mean.dist = mean(euclidean_played_dx_dy, na.rm = T)) %>% 
  ggplot(aes(x=bin, y=mean.dist, fill=isAlphaGo)) +
  geom_col(position="dodge") + 
  labs(x="Move number (binned into 10-percentiles)",
       y="Mean Euclidean distance from the previous move")
```




